Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-11-29 Manuscript/Manuscript.Rmd
Make full sentences from bullets in first part introduction. Edit and add explanation missingness figure.

# functions
miceadds::source.all(path = "../R")

@@

- What is clustering/multilevel data? In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering).  -> ADD: timeseries also in Discussion section.

- What do we mean by clustering? In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc. In other fields: e.g. official stats clustering at country-level, or social sciences clustering at school-level (related to the sampling design).

- What is heterogeneity? I.e. variability within studies vs. variability between studies

- What does multilevel data look like? ADD: figure to show difference between patient-level datapoints vs cluster-level datapoints. Maybe also add different data frame formats (or just explain in text that there's long and wide formats).

- What methods are required to analyze multilevel data?	Add references, e.g. @hox17 and @jong21. At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.

@@

- How can we categorize this? Systematic vs sporadic missingness, see @resc13. ADD: visualization of systematic vs sporadic missingness. Within systematic we have always missing (same value per cluster) and non-measured variables (may differ per patient). TODO: adjust md pattern to match text. -> syst may vary or same for all patients (observations/participants).

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(
    pat = rep("systematic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)),
  tibble(
    pat = rep("sporadic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = c("observed", "missing", rep("observed", 3)),
    X2 = c(rep("observed", 2), rep("missing", 3)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
) %>%
  mutate(across(everything(), as.factor),
         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE),
         text = rep(c(rep(c("1","", "", ""), 3), rep(c("2","", "", ""), 2)),2))
# plot
ggplot(patterns, aes(
  x = name,
  y = rownr,
  color = value,
  type = pat,
  width = 0.8, 
  height = 0.8,
  label = text
)) +
  geom_tile(fill = "white", size = 2) +
  geom_text() +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = 5:1) +
  scale_color_manual(values = plot_col, name = "Datapoint") +
  facet_wrap( ~ pat, ncol = 2, strip.position = "bottom") +
  theme_minimal() +
  theme(
    legend.position = "none", 
    panel.grid.major = element_blank()) +
  labs(x = "", y = "Observation (e.g., patient)")
# TODO: make it one plot with sporadic and systematic
# TODO: participant/response/*sample units*/records instead of obs on the y axis

@@

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data
pop <- readRDS("../Data/popNCR.RDS")
# ICCs
icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
```
- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.

@@

miss_col <- mice:::mdc(1:2) %>% setNames(c("FALSE", "TRUE"))

# x = "popular"
# z = "popteach"
plot_conditional <- function(data, x, z, cluster, ...){
  ggplot(data, aes(x = get(x), color = !is.na(get(z)))) + # group = get(cluster)
  geom_density(alpha = 0.01) +
  scale_color_manual(values = miss_col, name = z) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = x)
}
plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: pooteach obs/mis
# TODO: add title
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
```

@@

- Timeseries: and polynomial relationship in the clustering.

>>>>

# functions
miceadds::source.all(path = "../R")
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}

@@

We talk of multilevel data when there is some kind of hierarchy or clustering in a dataset. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by the sampling design. In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'. For reasons of brevity, we only discuss clustering between units, not longitudinal data (within-unit clustering).

Multilevel data requires special care when doing any sort of analysis. The cluster to which a unit belongs may influence the unit-level observations, and since clusters are made up of units, clusters depend on units as well [@hox17]. [Explain ICC here? The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC). The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster.] These relations can and should be taken into account when developing analysis models for multilevel data. At least, such models include a separate intercept term for each cluster. But there may also be random predictor effects and/or random error terms (residual error variances), see e.g. @hox17 and @jong21. Heterogeneity refers to variability within clusters vs. variability between clusters. There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

- How can we categorize this? Systematic vs sporadic missingness, see @resc13. Within systematic we have two flavors: unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters). In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest. Variable `X1` is completely observed. Variable `X2` is systematically missing, `X3` is sporadically missing. The unobserved value for units 4 and 5 on variable `X2` may be the same or different, defining which type of systematic missingness is happening.

```{r patterns, echo=FALSE}
# missingness indicator plot (monotone vs non-monotone)
miss_ind <- tibble(
    rownr = 5:1,
    unit = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c("missing", rep("observed", 3), "missing")
  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))

# add text 
miss_ind$text <- ""
miss_ind[miss_ind$name == "unit", "text"] <- as.character(1:5)
miss_ind[miss_ind$name == "cluster", "text"] <- as.character(c(1,1,1,2,2))

# plot
miss_ind %>% 
  mutate(name = factor(name, levels = unique(name))) %>% 
ggplot(aes(
  x = name,
  y = rownr,
  color = value,
  width = 0.8, 
  height = 0.8,
  label = text
)) +
  geom_tile(fill = "white", size = 2) +
  geom_text(show.legend = FALSE) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = 5:1) +
  scale_color_manual(values = plot_col, name = "") +
  theme_minimal() +
  theme(
    legend.position = "bottom", 
    panel.grid.major = element_blank()) +
  labs(x = "", y = "")
# TODO: make it one plot with sporadic and systematic
# TODO: participant/response/*sample units*/records instead of obs on the y axis

@@

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))

# ICCs
icc_pop <- multilevel::ICC1(aov(popular ~ as.factor(class), data = pop))
icc_teach <- multilevel::ICC1(aov(popteach ~ as.factor(class), data = pop))
```
- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.

@@

plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))

# x = "popular"
# z = "popteach"
plot_conditional <- function(data, x, z, cluster, ...){
  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
  geom_density() +
  scale_color_manual(values = plot_col, name = z) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = x)
}

plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: pooteach obs/mis
# TODO: add title
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
```

@@

- More complex data types: timeseries and polynomial relationship in the clustering.

### 2021-11-29 Manuscript/Manuscript.tex
Make full sentences from bullets in first part introduction. Edit and add explanation missingness figure.

\begin{itemize}
\item
  What is clustering/multilevel data? In this paper, we discuss grouped
  observations, not longitudinal data (within-patient clustering).
  -\textgreater{} ADD: timeseries also in Discussion section.
\item
  What do we mean by clustering? In the medical field: Clustering by
  studies (IPDMA), hospitals in registries, multi-center studies etc. In
  other fields: e.g.~official stats clustering at country-level, or
  social sciences clustering at school-level (related to the sampling
  design).
\item
  What is heterogeneity? I.e. variability within studies vs.~variability
  between studies
\item
  What does multilevel data look like? ADD: figure to show difference
  between patient-level datapoints vs cluster-level datapoints. Maybe
  also add different data frame formats (or just explain in text that
  there's long and wide formats).
\item
  What methods are required to analyze multilevel data? Add references,
  e.g. \citet{hox17} and \citet{jong21}. At least explain difference
  random effects for intercept term, predictor effects, and/or variance
  residual error.
\end{itemize}

@@

\begin{itemize}
\item
  Why/where does missingness occur in multilevel data? I.e., not only
  patient-level but also cluster-level.
\item
  How can we categorize this? Systematic vs sporadic missingness, see
  \citet{resc13}. ADD: visualization of systematic vs sporadic
  missingness. Within systematic we have always missing (same value per
  cluster) and non-measured variables (may differ per patient). TODO:
  adjust md pattern to match text. -\textgreater{} syst may vary or same
  for all patients (observations/participants).
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> pred 
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach
pupil        0     0      1   1    1       1        1
class        0     0      1   1    1       1        1
extrav       0     0      0   1    1       1        1
sex          0     0      1   0    1       1        1
texp         0     0      1   1    0       1        1
popular      0     0      1   1    1       0        1
popteach     0     0      1   1    1       1        0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
R> 
R> # check convergence of the imputation model
R> plot(imp_ignored)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  5.24      5   1  10.0  1.26
sex*     2000  1.51      2   1   2.0  0.50
texp     2000 12.63     13   2  25.0  6.27
popular  2000  5.01      5   0   9.1  1.37
popteach 2000  5.02      5   1  10.0  1.39
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_ignored)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- data.frame(
+   vars = c("popular", "popteach", "texp"), 
+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                multilevel::ICC1(aov(popteach ~ class, pop)),
+                multilevel::ICC1(aov(texp ~ class, pop))), 
+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+   )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2743092
2 popteach  0.3138658 0.2479930
3     texp  1.0000000 0.4510366
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach
pupil        0     1      1   1    1       1        1
class        0     0      1   1    1       1        1
extrav       0     1      0   1    1       1        1
sex          0     1      1   0    1       1        1
texp         0     1      1   1    0       1        1
popular      0     1      1   1    1       0        1
popteach     0     1      1   1    1       1        0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 190
\end{CodeOutput}
\begin{CodeInput}
R> # check logged events
R> head(imp_predictor$loggedEvents)
\end{CodeInput}
\begin{CodeOutput}
  it im      dep meth  out
1  1  1  popular  pmm texp
2  1  1 popteach  pmm texp
3  1  2  popular  pmm texp
4  1  2 popteach  pmm texp
5  1  3  popular  pmm texp
6  1  3 popteach  pmm texp
\end{CodeOutput}
\begin{CodeInput}
R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
R> 
R> # check convergence of the imputation model
R> plot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   2000  5.20    5.0   1  10.0  1.28
sex*     2000  1.52    2.0   1   2.0  0.50
texp     2000 14.26   15.0   2  25.0  6.55
popular  2000  5.05    5.1   0   9.1  1.38
popteach 2000  5.03    5.0   1  10.0  1.39
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- ICCs %>% cbind(
+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2743092 0.3694899
2 popteach  0.3138658 0.2479930 0.3335612
3     texp  1.0000000 0.4510366 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  JOMO in \texttt{mice} --\textgreater{} on the side for now
\item
  Additional levels of clustering
\item
  Timeseries: and polynomial relationship in the clustering.
\end{itemize}

>>>>

We talk of multilevel data when there is some kind of hierarchy or
clustering in a dataset. In the typical case, individuals are nested
within groups, but there are many different types of multilevel data. In
the medical field clustering occurs at e.g., the hospitals/center level
in registry data, or at the study-level in meta-analyses (IPDMA). In the
social sciences and official statistics we can find clustering e.g.~at
the country-level, or as imposed by the sampling design. In this paper,
we will refer to the grouping variable as `cluster', and the grouped
variable as `(sample) unit'. For reasons of brevity, we only discuss
clustering between units, not longitudinal data (within-unit
clustering).

Multilevel data requires special care when doing any sort of analysis.
The cluster to which a unit belongs may influence the unit-level
observations, and since clusters are made up of units, clusters depend
on units as well \citep{hox17}. {[}Explain ICC here? The percentage of
variance attributed to the cluster-level is expressed by the intra-class
coefficient (ICC). The ICC can also be interpreted as the expected
correlation between two randomly sampled units in same cluster.{]} These
relations can and should be taken into account when developing analysis
models for multilevel data. At least, such models include a separate
intercept term for each cluster. But there may also be random predictor
effects and/or random error terms (residual error variances), see e.g.
\citet{hox17} and \citet{jong21}. Heterogeneity refers to variability
within clusters vs.~variability between clusters. There are many names
for models that take clustering into account. Some popular examples are
`multilevel models', `hierarchical models', `mixed effect models' and
`random effect models'.

@@

\begin{itemize}
\item
  Why/where does missingness occur in multilevel data? I.e., not only
  patient-level but also cluster-level.
\item
  How can we categorize this? Systematic vs sporadic missingness, see
  \citet{resc13}. Within systematic we have two flavors: unobserved
  constants (same value per cluster) and non-measured random variables
  (which may differ per unit within clusters). In figure 1, we show a
  dataset with units in the rows and variables in the columns, there are
  5 units nested within 2 clusters, and 3 variables of interest.
  Variable \texttt{X1} is completely observed. Variable \texttt{X2} is
  systematically missing, \texttt{X3} is sporadically missing. The
  unobserved value for units 4 and 5 on variable \texttt{X2} may be the
  same or different, defining which type of systematic missingness is
  happening.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> pred 
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach school
pupil        0     0      1   1    1       1        1      1
class        0     0      1   1    1       1        1      1
extrav       0     0      0   1    1       1        1      1
sex          0     0      1   0    1       1        1      1
texp         0     0      1   1    0       1        1      1
popular      0     0      1   1    1       0        1      1
popteach     0     0      1   1    1       1        0      1
school       0     0      1   1    1       1        1      0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
R> 
R> # check convergence of the imputation model
R> plot(imp_ignored)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  5.25      5   1  10.0  1.27
sex*     2000  1.53      2   1   2.0  0.50
texp     2000 12.48     12   2  25.0  6.29
popular  2000  4.99      5   0   9.1  1.35
popteach 2000  5.02      5   1  10.0  1.39
school   2000  5.54      6   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
R> # TODO: pool mean median and sd
R> # feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
R> # TODO: add FMI for each of the estimates? at least for the mean
R> 
R> # further inspection of the imputations
R> densityplot(imp_ignored)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- data.frame(
+   vars = c("popular", "popteach", "texp"), 
+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                multilevel::ICC1(aov(popteach ~ class, pop)),
+                multilevel::ICC1(aov(texp ~ class, pop))), 
+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+   )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2716802
2 popteach  0.3138658 0.2528468
3     texp  1.0000000 0.4395296
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach school
pupil        0     1      1   1    1       1        1      1
class        0     0      1   1    1       1        1      1
extrav       0     1      0   1    1       1        1      1
sex          0     1      1   0    1       1        1      1
texp         0     1      1   1    0       1        1      1
popular      0     1      1   1    1       0        1      1
popteach     0     1      1   1    1       1        0      1
school       0     1      1   1    1       1        1      0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 325
\end{CodeOutput}
\begin{CodeInput}
R> # check logged events
R> head(imp_predictor$loggedEvents)
\end{CodeInput}
\begin{CodeOutput}
  it im      dep   meth
1  1  1   extrav    pmm
2  1  1      sex logreg
3  1  1     texp    pmm
4  1  1  popular    pmm
5  1  1 popteach    pmm
6  1  1 popteach    pmm
                                                                                                                                                                                                                                                       out
1                                                                                                                                                                                                                                                   school
2                                                                                                                                                                                                                                                   school
3                                                                                                                                                                                                                                                   school
4                                                                                                                                                                                                                                             texp, school
5                                                                                                                                                                                                                                                     texp
6 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
\end{CodeOutput}
\begin{CodeInput}
R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
R> 
R> # check convergence of the imputation model
R> plot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  6.20      6   1  10.0  2.00
sex*     2000  1.52      2   1   2.0  0.50
texp     2000 14.26     15   2  25.0  6.55
popular  2000  5.03      5   0   9.1  1.35
popteach 2000  5.03      5   1  10.0  1.38
school   2000  5.54      6   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- ICCs %>% cbind(
+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2716802 0.3550567
2 popteach  0.3138658 0.2528468 0.3306035
3     texp  1.0000000 0.4395296 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  JOMO in \texttt{mice} --\textgreater{} on the side for now
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\end{itemize}
